---
title: "Hypothesis Testing"
author: "Chapter 4, Lab 3: Template"
date: "OpenIntro Biostatistics"

fontsize: 11pt
geometry: margin=1in
output: pdf_document
---

The previous labs in this unit discussed the calculation of point estimates and interval estimates for a population mean. An interval estimate produces a range of plausible values for $\mu$ based on the observed data.

Hypothesis testing operates under a different perspective---under the working hypothesis that the population mean is a particular value $\mu_0$, what is the likelihood of observing a sample with mean $\overline{x}$? If there is a very small chance of observing a sample with mean $\overline{x}$ when $\mu = \mu_0$, this represents evidence against the working hypothesis and suggests that the $\mu$ does not equal $\mu_0$.

The material in this lab corresponds to Section 4.3.1 - 4.3.2 of *OpenIntro Biostatistics*.

####Introduction to hypothesis testing

This lab uses data from the National Health and Nutrition Examination Survey (NHANES), a survey conducted annually by the US Centers for Disease Control (CDC).\footnote{The dataset was first introduced in Chapter 1, Lab 1 (Introduction to Data).} The complete \texttt{NHANES} dataset contains 10,000 observations, which will be the artificial target population. 

\begin{center}
\textit{Are American adults getting enough sleep?}
\end{center}

A 2016 study from the CDC reported that more than a third of American adults are not getting enough sleep on a regular basis.\footnote{\url{https://www.cdc.gov/media/releases/2016/p0215-enough-sleep.html}} The National Sleep Foundation recommends that adults need between 7 to 9 hours of sleep per night to function well. Consistent sleep deprivation is known to increase risk of health problems, negatively affect cognitive processes, and contribute to depressive symptoms. 

The dataset \texttt{nhanes.samp.adult} in the \texttt{oibiostat} package contains sleep information for 135 participants ages 21 years or older that were randomly sampled from the \texttt{NHANES} population. The variable \texttt{SleepHrsNight} contains the self-reported number of hours a participant usually gets at night on weekdays or workdays. 

1. Explore the distribution of \texttt{SleepHrsNight} in \texttt{nhanes.samp.adult}.

  a) Using numerical and graphical summaries, describe the distribution of nightly sleep hours in \texttt{nhanes.samp.adult}.
    
```{r}
#load the data

#numerical summaries

#graphical summaries

```



    
  b) Based on the distribution of nightly sleep hours in the sample, does it seem like the population mean nightly sleep hours will be outside the range defined as adequate (7 - 9 hours)?





    
2. Calculate a 95\% confidence interval for nightly sleep hours using \texttt{nhanes.samp.adult} and interpret the interval. 

<!---

For example, the command to calculate a 95\% confidence interval for \texttt{height} is

```{r, eval = FALSE}
t.test(nhanes.samp.adult$Height, na.rm = TRUE, conf.level = 0.95)$conf.int
```

--->

```{r}
#calculate a confidence interval

```




3. Conduct a hypothesis test to assess whether on average, American adults are getting enough sleep. Let $\mu_0$ be 8 hours, the midpoint of the range defined as adequate.

  a) Formulate null and alternative hypotheses. The symbol $\mu$ denotes the population mean, while $\mu_0$ refers to the numeric value specified by the null hypothesis. 
    
    
    
    
    
  b) Specify a significance level, $\alpha$.
    
    
    
    
    
    
  c) Calculate the test statistic.
    
    \[t = \frac{\overline{x} - \mu_0}{s/\sqrt{n}} \]
    
    
```{r, eval = FALSE}
#use r as a calculator
x.bar = 
mu.0 = 
s =
n = length(nhanes.samp.adult$SleepHrsNight)

t = (x.bar - mu.0)/(s/sqrt(n))
t
```

    
  d) Calculate the $p$-value.

<!---

To calculate a tail area from a t-statistic, use the pt() function, which is analogous to pnorm(). Specify the t-statistic and the degrees of freedom; set lower.tail = FALSE if you want an upper tail area (i.e., area to the right of the t-statistic).

--->

```{r, eval = FALSE}
pt( , df = , lower.tail = )
```



    
  e) Draw a conclusion.
    




    
####Hypothesis Testing: Cholesterol Level

High cholesterol is a major controllable risk factor for coronary heart disease, heart attack, and stroke. According to general guidelines, a total cholesterol level below 5.2 mmol/L is desirable, in the 5.2 - 6.2 mmol/L range is borderline high, and above 6.2 mmol/L is high. 

4. Describe the distribution of total cholesterol in \texttt{nhanes.samp.adult}. Does it seem that most individuals have a cholesterol level considered desirable?



```{r}

```




5. Conduct a hypothesis test to assess whether mean total cholesterol in the \texttt{NHANES} "population" is equal to 5.2 mmol/L, using the data in \texttt{nhanes.samp.adult}.

  a) Choose whether to conduct a one-sided or two-sided test. Formulate null and alternative hypotheses.
  
  
  
  
  
  b) Specify a significance level, $\alpha$.
  
  
  
  
  
  c) Calculate the test statistic. (Note: Be careful of missing data values.)
  
  
  
```{r, eval = FALSE}
#use r as a calculator
x.bar = 
mu.0 = 
s =
n = 

t = (x.bar - mu.0)/(s/sqrt(n))
t
```  
  
  
  d) Calculate the $p$-value.
  
   

```{r}

```

  
  
  e) Confirm your calculations in parts c) and d) using \texttt{t.test()}.
  
<!---
To use the t.test() command...

x is the name of the variable containing the sample data
mu is the mu_0 value specified by the null and alternative hypotheses
alternative can be set to either
  "two.sided" for a two-sided test
  "greater" for a one-sided test, mu > mu_0
  "less" for a one-sided test, mu < mu_0
--->

```{r, eval = FALSE}
t.test(x = , mu = , alternative = )
```

    
  
  f) Draw a conclusion.




####Hypothesis Testing: Body Temperature

Mean body temperature is commonly accepted to be 98.6 degrees Fahrenheit. The origin of this benchmark value is credited to Carl Wunderlich, a scientist working in clinical thermometry in the late 19th century. A study was conducted in 1992 to evaluate whether population mean body temperature among healthy adults is really 98.6 F. Data were collected from healthy volunteers who had agreed to participate in a separate set of vaccine trials; these data are in the dataset \texttt{thermometry} in the \texttt{oibiostat} package.

6. Conduct a formal hypothesis test to evaluate if mean body temperature is really 98.6 F. Be sure to clearly specify the hypotheses and report the conclusions.




```{r}
#load the data

#calculate the test statistic and p-value

```





7. Calculate a 99\% confidence interval for mean body temperature in healthy adults. Does the interval contain 98.6 F?

```{r}
#calculate a confidence interval


```





8. Briefly summarize the conclusions of this analysis in language accessible to a general audience. What is a possible reason for the observed discrepancy between the 1992 data and Wunderlich's data?




