---
title: "Inference for Two-Way Tables"
author: "Chapter 8, Lab 2: Template"
date: "OpenIntro Biostatistics"

fontsize: 11pt
geometry: margin=1in
output: pdf_document
---

\begin{small}
	
	\textbf{Topics}
	\begin{itemize}
	  \item The $\chi^2$ test for independence
	  \item Measures of association in two-by-two tables
	\end{itemize}
	
\end{small}

This lab generalizes inference for binomial proportions to the setting of two-way contingency tables. Hypothesis testing in a two-way table assesses whether the two variables of interest are associated; this approach can be applied to settings with two or more groups and for responses that have two or more categories. Measures of association in two-by-two tables are also discussed.

The material in this lab corresponds to Sections 8.3 and 8.5 in *OpenIntro Biostatistics*.

### Introduction

*The $\chi^2$ test of independence*

In the $\chi^2$ test of independence, the observed number of cell counts are compared to the number of **expected** cell counts, where the expected counts are calculated under the null hypothesis.

  - $H_0$: the row and column variables are not associated

  - $H_A$: the row and column variables are associated

The expected count for the $i^{th}$ row and $j^{th}$ column is
\[E_{i, j} = \dfrac{(\text{row $i$ total}) \times (\text{column $j$ total}) }{n}, \]
where $n$ is the total number of observations.

Assumptions for the $\chi^2$ test:

  - *Independence*. Each case that contributes a count to the table must be independent of all other cases in the table.
  
  - *Sample size*. Each expected cell count must be greater than or equal to 10. For tables larger than $2 \times 2$, it is appropriate to use the test if no more than 1/5 of the expected counts are less than 5, and all expected counts are greater than 1.

The **$\chi^2$ test statistic** is calculated as
\[\chi^2 = \sum_{i = 1}^r \sum_{j = 1}^c \dfrac{(O_{i, j} - E_{i, j})^2}{E_{i, j}}, \]
and is approximately distributed $\chi^2$ with degrees of freedom $(r - 1)(c - 1)$, where $r$ is the number of rows and $c$ is the number of columns. $O_{i, j}$ represents the observed count in row $i$, column $j$.

For each cell in a table, the **residual** equals
\[\dfrac{O_{i, j} - E_{i, j}}{\sqrt{E_{i,j}}}. \]
Residuals with a large magnitude contribute the most to the $\chi^2$ statistic. If a residual is positive, the observed value is greater than the expected value; if a residual is negative, the observed value is less than the expected.

*Measures of association in two-by-two tables*

Chapter 1 introduced the **relative risk (RR)**, a measure of the risk of a certain event occurring in one group relative to the risk of the event occurring in another group, as a numerical summary for two-by-two ($2 \times 2$) tables. The relative risk can also be thought of as a measure of association.

Consider the following hypothetical two-by-two table. The relative risk of Outcome A can be calculated by using either Group 1 or Group 2 as the reference group:

\begin{table}[h!]
	\centering
	\begin{tabular}{r|rrr}
		\hline
		& Outcome A & Outcome B & Sum\\ 
		\hline
		Group 1 & $a$ & $b$ & $a + b$ \\ 
		Group 2 & $c$ & $d$ & $c + d$ \\
		Sum & $a + c$ & $b + d$ & $a + b + c + d = n$ \\
		\hline
	\end{tabular}	
	\caption{A hypothetical two-by-two table of outcome by group.}
\end{table}


\[RR_{\text{A, comparing Group 1 to Group 2}} = \dfrac{a/(a + b)}{c/(c+d)} \]
\[RR_{\text{A, comparing Group 2 to Group 1}} = \dfrac{c/(c + d)}{a/(a+b)} \]

The relative risk is only valid for tables where the proportions $a/(a + b)$ and $c/(c + d)$ represent the incidence of Outcome A within the populations from which Groups 1 and 2 are sampled.

The **odds ratio (OR)** is a measure of association that remains applicable even when it is not possible to estimate incidence of an outcome from the sample data. The **odds** of Outcome A in Group 1 are $a/b$, while the odds of Outcome A in Group 2 are $c/d$.

\[OR_{\text{A, comparing Group 1 to Group 2}} = \dfrac{a/b}{c/d} = \dfrac{ad}{bc} \]
\[OR_{\text{A, comparing Group 2 to Group 1}} = \dfrac{c/d}{a/b} = \dfrac{bc}{ad} \]

\newpage

### The $\chi^2$ test of independence

1. In resource-limited settings, single-dose nevirapine (NVP) is given to an HIV-positive woman during birth to prevent mother-to-child transmission of the virus. Exposure of the infant to NVP may foster the growth of more virulent strains of the virus in the child.

    If a child is HIV-positive, should they be treated with NVP or a more expensive drug, lopinarvir (LPV)? In this setting, success means preventing a growth of the virus in the child (i.e., preventing virologic failure).
    
    \begin{center}
\begin{tabular}{l|cc|c} 
   & \textbf{NVP} & \textbf{LPV} & \textbf{Total}\\ \hline
  \textbf{Virologic Failure} & 60 & 27 & 87  \\
  \textbf{Stable Disease} & 87 & 113 & 200 \\ \hline
  \textbf{Total} & 147 & 140 & 287  \\ 
\end{tabular}\\
\end{center}

    \vspace{0.5cm}
    
a) State the null and alternative hypotheses.




    
b) Calculate the expected cell counts.


```{r, eval = FALSE}
#use r as a calculator

#set parameters
n = 287
row.1.total = 87
row.2.total = 200
col.1.total = 147
col.2.total = 140

#calculate expected values
exp.1.1 = (row.1.total * col.1.total)/n
exp.1.1

exp.1.2 = 
  
exp.2.1 =
  
exp.2.2 =
```

    
c) Check the assumptions for using the $\chi^2$ test.




    
d) Calculate the $\chi^2$ test statistic.
    

```{r, eval = FALSE}
#use r as a calculator
obs.1.1 = 60
chi.sq.1.1 = ((obs.1.1 - exp.1.1)^2)/exp.1.1

obs.1.2 =
chi.sq.1.2 =   

obs.2.1 =
chi.sq.2.1 =
  
obs.2.2 =
chi.sq.2.2 =
  
chi.sq = chi.sq.1.1 + chi.sq.1.2 + chi.sq.2.1 + chi.sq.2.2
chi.sq
```



e) Calculate the $p$-value for the test statistic using \texttt{pchisq()}. The $p$-value represents the probability of observing a result as or more extreme than the sample data.


```{r, eval = FALSE}
#use pchisq()
pchisq( , df = )
```

    
f) Confirm the results from parts c) and d) using \texttt{chisq.test()}. Note that the value of the test statistic will be slightly different because \textsf{R} is applying a 'continuity correction'.

```{r}
#enter the data as a table
hiv.table = matrix(c(60, 27, 87, 113), 
                  nrow = 2, ncol = 2, byrow = T)

#use chisq.test()
chisq.test(hiv.table)
```

    
g) Summarize the conclusions; be sure to include which drug is recommended for treatment, based on the data.



```{r}
#look at residuals
chisq.test(hiv.table)$resid
```

    
h) Repeat the analysis using inference for the difference of two proportions and confirm that the results are the same.


```{r}
#use prop.test( )

```

    
2. In the PREVEND study introduced in Chapter 6, researchers measured various features of study participants, including data on statin use and highest level of education attained. From the data in \texttt{prevend.samp}, is there evidence of an association between statin use and educational level? Summarize the results.



```{r}
#load the data
library(oibiostat)
data("prevend.samp")

#convert variables to factors
prevend.samp$Statin = factor(prevend.samp$Statin, levels = c(0, 1),
                             labels = c("NonUser", "User"))

prevend.samp$Education = factor(prevend.samp$Education, levels = 0:3,
                                labels = c("Primary", "LowerSec",
                                           "UpperSec", "Univ"))

#create a table

#run chi-squared test

```


\newpage

### Measures of association in two-by-two tables

3. Suppose a study is conducted to assess the association between smoking and cardiovascular disease (CVD). Researchers recruited a group of 231 study participants then categorized them according to smoking and disease status: 111 are smokers, while 40 smokers and 32 non-smokers have CVD. Calculate and interpret the relative risk of CVD.



```{r}
#use r as a calculator

```


4. Suppose another study is conducted to assess the association between smoking and CVD, but researchers use a different design: 90 individuals with CVD and 110 individuals without CVD are recruited. 40 of the individuals with CVD are smokers, and 80 of the individuals without CVD are non-smokers.


a) Is relative risk an appropriate measure of association for these data? Explain your answer.
    


b) Calculate the odds of CVD among smokers and the odds of CVD among non-smokers.
    

```{r}
#use r as a calculator

```



c) Calculate and interpret the odds ratio of CVD, comparing smokers to non-smokers.
    

```{r}
#use r as a calculator
```


d) What would an odds ratio of CVD (comparing smokers to non-smokers) equal to 1 represent, in terms of the association between smoking and CVD? What would an odds ratio of CVD less than 1 represent?



