---
title: "Lab Notes"
author: "Chapter 9"
date: "OpenIntro Biostatistics"

fontsize: 11pt
geometry: margin=1in

output:
  pdf_document:
    includes:
      in_header: ../../header.tex
    fig_width: 6.5
    fig_height: 3.5
---

# Overview

1. Simple Logistic Regression
    - *OI Biostat* Section 9.xx
    
2. Multiple Logistic Regression
    - *OI Biostat* Section 9.xx

Lab 1 inroduces simple logistic regression, a model for the association of a binary response variable with a single predictor variable.

Lab 2 discusses multiple logistic regression, an extension of simple logistic regression that allows for several predictors. The use of the Akaike Information Criterion as a metric for model selection is also discussed.

\newpage


# Lab 1: Simple Logistic Regression

### Fitting a Logistic Regression Model

The \textbf{\texttt{glm()}} function is used to fit logistic regression models. It has the following generic structure:

```{r, eval = FALSE}
glm(y ~ x, data, family = binomial(link = "logit"))
```

where the first argument specifies the variables used in the model; in this example, the model regresses a response variable \texttt{y} against an explanatory variable \texttt{x}. The second argument is used only when the dataframe name is not already specified in the first argument. Running the function creates an *object* (of class '\texttt{lm}' and '\texttt{glm}') that contains several components, such as the model coefficients. The model coefficients are directly displayed upon running \texttt{glm()}, while other components can be accessed through either the \texttt{\$} notation or specific functions like \texttt{summary()}. The argument \texttt{family = binomial(link = "logit")} is specific to logistic regression; the texttt{glm()} function is capable of running families of general linear models that are not discussed in this course.

The following example shows fitting a linear model that predicts the estimated log odds of death before discharge from resting heart rate, using data from \texttt{icu}.

```{r}
#load the data
library(aplore3)
data("icu")

#fitting logistic model
glm(sta ~ hra, data = icu, family = binomial(link = "logit"))
```

To fit a linear model that predicts the estimated log odds of survival to discharge from resting heart rate, it is necessary to relevel the factor \texttt{sta} such that a \texttt{1} corresponds to individuals who survived to discharge. This can be accomplished with \textbf{\texttt{factor()}} and \textbf{\texttt{rev()}}. The \texttt{rev()} function reverses elements. In the example below, applying \texttt{rev)} to a vector {1, 2, 3} produces a vector {3, 2, 1}.

```{r}
#check levels
levels(icu$sta)

#relevel survival
icu$sta = factor(icu$sta, levels = rev(levels(icu$sta)))

#check levels
levels(icu$sta)

#example of using rev()
a = c(1, 2, 3)
rev(a)
```

The following example shows outputting the model summary, selectively outputting model coefficients from the model fit, and extracting the numeric value of a coefficient.

```{r}
#name the model
model.hra = glm(sta ~ hra, data = icu, family = binomial(link = "logit"))

#model summary
summary(model.hra)

#model summary of coefficients
summary(model.hra)$coef

#extract value of slope coefficient
coef(model.hra)[2]
```

As in linear regression, the \texttt{predict()} function can be used to evaluate the regression equation for specific values of a predictor variable. The following example shows predicting the estimated log odds of survival to discharge for an individual with resting heart rate of 98 bpm.

```{r}
predict(model.hra, newdata = data.frame(hra = 98))
```


\newpage

# Lab 2: Multiple Logistic Regression

### Working with Several Predictors

The \textbf{\texttt{glm()}} function is used to fit linear models. It has the following generic structure:

```{r, eval = FALSE}
glm(y ~ x1 + x2, data, family = binomial(link = "logit"))
```

where the first argument specifies the variables used in the model; in this example, the model regresses a response variable \texttt{y} against two explanatory variables \texttt{x1} and \texttt{x2}. Additional predictor variables can be added to the model formula with the \texttt{+} symbol, and an interaction between two variables is specified with the \texttt{*} symbol.

The following example shows fitting a linear model that predicts the estimated log odds of survival to discharge from age and gender, and a linear model that predicts the estimated log odds of survival to discharge from age, gender, and their interaction.

```{r}
#fitting model with age and gender
glm(sta ~ age + gender, data = icu, family = binomial(link = "logit"))

#fitting model with age, gender, and an interaction term
glm(sta ~ age*gender, data = icu, family = binomial(link = "logit"))
```

### Calculating AIC

The AIC of a logistic model can be extracted from \texttt{summary()} or computed via the \textbf{\texttt{AIC()}} function.

The following example shows how to output the AIC from the model predicting estimated odds of survival to discharge from resting heart rate.

```{r}
#use summary()$aic
summary(model.hra)$aic

#use AIC()
AIC(model.hra)
```


### Collapsing Factor Levels

The \texttt{factor()} function can also be used to collapse levels of a factor. 

The following example shows the re-defining of the levels of \texttt{loc}; the variable initially has three levels (\texttt{Nothing}, \texttt{Stupor}, and \texttt{Coma}). The levels \texttt{Stupor} and \texttt{Coma} can be combined into a single level \texttt{Unconscious}, while the level \texttt{Nothing} is renamed \texttt{Conscious}. 

```{r}
#view levels of loc
levels(icu$loc)

#create the loc.binary variable
icu$loc.binary = icu$loc

#redefine the factor levels of loc.binary
levels(icu$loc.binary) = list("Conscious" = "Nothing", 
                              "Unconscious" = c("Stupor", "Coma"))

#view levels of loc.binary
levels(icu$loc.binary)

#compare tables
table(icu$loc); table(icu$loc.binary)
```

