---
title: "Introduction to Random Variables"
author: "Chapter 3, Lab 1: Template"
date: "OpenIntro Biostatistics"

fontsize: 11pt
geometry: margin=1in
output: pdf_document
---

\begin{small}
	
	\textbf{Topics}
	\begin{itemize}
	  \item Distributions of random variables
	  \item Mean and variance of a random variable
	  \item Binomial distribution
	\end{itemize}
	
\end{small}

A random variable numerically summarizes the possible outcomes of a random experiment. Formally, a random variable assigns numerical values to the outcomes of a random phenomenon. This lab introduces random variables by exploring the clinical trial example described at the beginning of Chapter 3 and by discussing the binomial distribution.

The material in this lab corresponds to Sections 3.1 and 3.2 of *OpenIntro Biostatistics*.

#### Clinical Trial Simulation

1. Suppose that a clinical trial to test a new drug will be conducted on 8 patients, in which the probability of a good response to the drug is thought to be 0.15. The following code simulates the trial with 500 replicates.

<!---
Set eval = TRUE for the results to display in the knitted PDF.
--->

```{r, eval = FALSE}
#define parameters
number.patients = 8
response.prob = 0.15
number.replicates = 500

#create empty vectors to store results
number.responses.replicate = vector("numeric", number.patients)
number.responses = vector("numeric", number.replicates)

#set the seed for a pseudo-random sample
set.seed(2018)

#simulate the trials
for(k in 1:number.replicates){
  
  number.responses.replicate = sample(c(0,1), size = number.patients,
                                      prob = c(1 - response.prob, response.prob),
                                      replace = TRUE)
  
  number.responses[k] = sum(number.responses.replicate)
  
}

#view the results

```

a) Run the simulation and view the results (*Hint*: make a plot). Describe the distribution of the number of good responses to the drug.
    



    

b) What value(s) for response probability would produce a left-skewed distribution? What value(s) would produce a symmetric distribution?





c) Based on the results of the simulation, estimate the probability that 0 patients respond well to the new drug. 
    
    
    
    

d) Let $X$ be a random variable defined as the number of patients who respond well to the experimental drug. Based on the results of the simulation, $\dots$

    i. Construct a probability distribution for $X$.
          
          
          
          
  
    ii. Calculate $E(X)$, where 
    \[E(X) 	= x_1 P(X=x_1) + \cdots + x_k P(X=x_k) = \sum_{i=1}^{k}x_iP(X=x_i)\]
  
  
  
  
  
    iii. Calculate $\text{Var}(X)$, where
    \[\text{Var}(X) = (x_1-\mu)^2 P(X=x_1) + \cdots + (x_k-\mu)^2 P(X=x_k) = \sum_{j=1}^{k} (x_j - \mu)^2 P(X=x_j) \]





\newpage

#### Binomial Distribution

Let $X$ represent the number of successes in a series of independent Bernoulli trials. Suppose the probability of a single trial being a success is $p$. The probability of observing exactly $k$ successes in $n$ independent trials is given by
\[P(X = k) = { n \choose k} p^k (1-p)^{n-k} = \frac{n!}{k!(n-k)!} p^k (1-p)^{n-k}. \]

For a binomial random variable, $E(X) = np$ and $\text{Var}(X) = np(1-p)$.

Binomial probabilities are calculated in \textsf{R} with the use of \texttt{dbinom} and \texttt{pbinom}.

The following code shows how to calculate $P(X = 5)$, $P(X \leq 5)$, and $P(X > 5)$ for $X \sim \text{Bin}(10, 0.35)$. 

```{r}
#probability X equals 5
dbinom(5, 10, 0.35)

#probability X is less than or equal to 5
pbinom(5, 10, 0.35)

#probability X is greater than 5
pbinom(5, 10, 0.35, lower.tail = FALSE) 
```


2. Why is the value returned by \texttt{pbinom(5, 10, 0.35)} greater than the value from \texttt{dbinom(5, 10, 0.35)}?






3. The hypothetical clinical trial discussed in Question 1 can be modeled with a binomial distribution. 

    a) Confirm that the clinical trial satisfies the conditions for a binomial experiment.
  
  
  
  
  
    b) Calculate $E(X)$ and $\text{Var}(X)$ using the formulas specific to the binomial distribution; compare the results to the answers from part d) of Question 1. Explain any observed discrepancies.





\newpage

4. Approximately 12,500 stocks of *Drosophila melanogaster* flies are kept at the Bloomingdale *Drosophila* Stock Center for research purposes. A 2006 study examined how many stocks were infected with Wolbachia, an intracellular microbe that can manipulate host reproduction for its own benefit. About 30\% of stocks were identified as infected. Researchers working with infected stocks should be cautious of the potential confounding effects that Wolbachia infection may have on experiments.

    Consider a random sample of 250 stocks. Let $X$ represent the number of infected stocks in the sample.
    
    a) Calculate the probability that exactly 60 stocks are infected.
    
    
```{r}

```

    
    
    b) Calculate the probability that at most 60 stocks are infected.
    
    
```{r}

```
    
    
    c) Calculate the probability that at least 80 stocks are infected. 
    
    
```{r}

```

    
    
    d) Assume that a researcher will use all of the stocks sampled for an experiment.  If the researcher wants to be sure that no more than 40\% of the stocks used for an experiment are infected, does it seem reasonable to take a random sample of 250 stocks?
    
    
```{r}

```

        
    
    e) Demonstrate how parts a) through d) can be approached through simulation.

```{r}
#define parameters

#create empty vectors to store results

#set the seed for a pseudo-random sample

#simulate the samples

#view the results

```

    
5. The National Vaccine Information Center estimates that 90\% of Americans have had chickenpox by the time they reach adulthood.

    a) Calculate the probability that exactly 97 out of 100 randomly sampled American adults had chickenpox during childhood.


```{r}

```


    b) Calculate the probability that exactly 3 out of a new sample of 100 American adults have not had chickenpox in their childhood.


```{r}

```

        
    c) What is the probability that at least 1 out of 10 randomly sampled American adult has had chickenpox?


```{r}

```


    d) What is the probability that at most 3 out of 10 randomly sampled American adults have not had chickenpox?


```{r}

```


    e) Suppose that the adults were sampled from the same city. Would it be appropriate to use the binomial distribution to calculate the probabilities from parts a) through d)?
  